MS_Intro_Short

### 2023-08-15 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
working way through frameworks

We considered prior literature on modelling practice, our own professional experience and the needs of our youth mental health modelling project, to identify core ethical responsibilities of CHEM developers, attributes of CHEMs that support fulfilment of these responsibilities and criteria against which these attributes can be assessed.

## Ethical responsibilities of CHEM developers 
Health economists have ethical responsibilities relating to the social acceptability, adequacy for purpose and beneficial impact of CHEMs.

@@

##	Sssessing ethical CHEM implementations
The acceptability, adequacy and public benefit responsibilities of health economic modellers are easier to state than to measure.  We have therefore focused our assessment criteria for ethical modelling practice on measurable attributes of computational model implementations. As described in Table \@ref(tab:timelygls),  we believe that implementing CHEMs that are transparent, reusable and updatable (TRU) can enable modellers to meet their ethical obligations. We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

@@

**Transparency** has been  as a core identified as a core feature of ethical public health modelling practice [@10.3389/fpubh.2017.00068]. Guidance on transparency in health economic modelling recommended that model code and data should be clearly documented, potentially with different versions for technical and non-technical users [@Eddy2012]. Notably, the same guidelines, published over ten years ago, did not include recommendations on sharing model code and data. However, more recent guidance recommends publicly dissemination of healthcare model artefacts using online repository services [@Erdemir2020]. Repositories such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. An essential component of quality assuring health economic models is verification - ensuring that calculations are correct and consistent with model specifications [@techver2019]. The extensiveness of verification checks in models implemented as software projects can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape]. This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

The **transparency** of a CHEM implementation can be assessed against the criteria:

-	T1: All model code, non-confidential data and testing procedures and outcomes are available in open access repositories.

-	T2: It is easy to see who developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.

Model **reusability** is the second attribute that we identify as being core to ethical CHEM implementations. Making a CHEM’s code, data and documentation publicly available is increasingly considered good practice, yet is not sufficient to promote model re-use.  Writing model algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task), is good scientific computing practice [@Wilson_2017] and facilitates model reuse without adaptation (generalisability). To support reuse with adaptation (transferability), computational implementations can store model code and data in distinct files and locations (as opposed to embedding data such as parameter values into source code).  Transferability can also be advanced through the use of object-oriented programming approaches [@ready4oop2022] that take advantage of the concept of inheritance [@8717448] to allow models to be first duplicated and then selectively modified. Generalisability and transferability are further enhanced if computational implementations adopt a modular approach, in which a model is constructed from multiple independently reusable and replaceable sub-models (modules) [@pan2021modular]. To grant permissions to others to use, test and adapt models and their components, health economists can avail of two broad categories of open source licensing options. Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which content can be re-used. An alternative approach is to use copyleft licenses [@copyleft2022] that can require content users to distribute any derivative works they create under similar open-source arrangements. 
<!-- Compared to using commercial modelling software, authoring CHEMs in an open-source language like R [@RCORE2022] makes it easier to -->
<!-- Using open-source development platforms and licenses can aid both generalizability and transferability. -->
<!-- The choices that CHEM developers make about model implementation and licensing will also shape who can use a model and for what purposes.  -->
<!-- are relatively easier to selectively modify. -->
The **reusability** of a CHEM implementation can be assessed against the criteria:

-	R1: Model code and data are implemented to facilitate both generalisability and transferability.

-	R2: Terms of use allow anyone to reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.

Models should be **updatable** so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change [@Jenkins2021; @@garcia2021cost]. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defense against model validity decay [@calder2018computational]. Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028]. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of Continuous Integration [@CI2017].  The risk of model revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated model code and data.
<!-- Currently, the core development team for a CHEM will be typically be funded to produce a project end-point deliverable whose specifications are well defined early in the project. For more complex and multi-purpose CHEMs, particularly those designed to be incorporated into decision support systems, it may be better for development teams to adopt Agile Software Development [@beck2001manifesto], an approach that has been recommended for complex public health software projects [@10.3389/fpubh.2022.899874]. An Agile model will be less clearly specified in the initial project plan, but will instead continually develop in response to the requirements and feedback of users, who are provided with an initial, simplified working version of the model at the earliest feasible opportunity. -->

The **updatability** of a CHEM implementation can be assessed against the criteria:

-	U1: Technical infrastructure is in place to support sustained development, testing, maintenance and version control of a model in collaboration with model users.

-	U2: Each new release of a model is retested, with changes implemented to minimize disruptions for existing model users.

>>>>

We considered prior literature on modelling practice, our own professional experience and the needs of modelling projects we are undertaking to identify some core ethical responsibilities of CHEM developers, attributes of CHEMs that can support fulfillment of these responsibilities and criteria against which these attributes can be assessed.

### Ethical responsibilities of CHEM developers 
Health economists have ethical responsibilities relating to the social acceptability, adequacy for purpose and beneficial impact of CHEMs.

@@

### CHEM attributes associated with ethical modelling practice
The acceptability, adequacy and public benefit responsibilities of health economic modellers are easier to state than to measure.  We have therefore focused our assessment criteria for ethical modelling practice on measurable attributes of computational model implementations. As described in Table \@ref(tab:timelygls),  we believe that implementing CHEMs that are transparent, reusable and updatable (TRU) can enable modellers to meet their ethical obligations.  We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

@@

**Transparency** has been identified as a core feature of ethical public health modelling practice [@10.3389/fpubh.2017.00068]. Guidance on transparency in health economic modelling recommended that model code and data should be clearly documented, potentially with different versions for technical and non-technical users [@Eddy2012]. Notably, the same guidelines, published over ten years ago, did not include recommendations on sharing model code and data. However, more recent guidance recommends publicly dissemination of healthcare model artefacts using online repository services [@Erdemir2020]. Repositories such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. An essential component of quality assuring health economic models is verification - ensuring that calculations are correct and consistent with model specifications [@techver2019]. The extensiveness of verification checks in models implemented as software projects can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape]. This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

Model **reusability** is the second attribute that we identify as being core to ethical CHEM implementations. Making a CHEM’s code, data and documentation publicly available is increasingly considered good practice, yet is not sufficient to promote model re-use.  Writing model algorithms as collections of functions (short, self-contained and reusable software routines that each perform a discrete task), is good scientific computing practice [@Wilson_2017] and facilitates model reuse without adaptation (generalisability). To support reuse with adaptation (transferability), computational implementations can store model code and data in distinct files and locations (as opposed to embedding data such as parameter values into source code).  Transferability can also be advanced through the use of object-oriented programming approaches [@ready4oop2022] that take advantage of the concept of inheritance [@8717448] to allow models to be first duplicated and then selectively modified. Generalisability and transferability are further enhanced if computational implementations adopt a modular approach, in which a model is constructed from multiple independently reusable and replaceable sub-models (modules) [@pan2021modular]. To grant permissions to others to use, test and adapt models and their components, health economists can avail of two broad categories of open source licensing options. Some guidance strongly recommends the use of permissive licensing [@Wilson_2017] that provides users with great flexibility as to the purposes (including commercial) for which content can be re-used. An alternative approach is to use copyleft licenses [@copyleft2022] that can require content users to distribute any derivative works they create under similar open-source arrangements. 
<!-- Compared to using commercial modelling software, authoring CHEMs in an open-source language like R [@RCORE2022] makes it easier to -->
<!-- Using open-source development platforms and licenses can aid both generalizability and transferability. -->
<!-- The choices that CHEM developers make about model implementation and licensing will also shape who can use a model and for what purposes.  -->
<!-- are relatively easier to selectively modify. -->

Models should be **updatable** so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change [@Jenkins2021; @garcia2021cost]. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defense against model validity decay [@calder2018computational]. Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028]. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of Continuous Integration [@CI2017].  The risk of model revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated model code and data.
<!-- Currently, the core development team for a CHEM will be typically be funded to produce a project end-point deliverable whose specifications are well defined early in the project. For more complex and multi-purpose CHEMs, particularly those designed to be incorporated into decision support systems, it may be better for development teams to adopt Agile Software Development [@beck2001manifesto], an approach that has been recommended for complex public health software projects [@10.3389/fpubh.2022.899874]. An Agile model will be less clearly specified in the initial project plan, but will instead continually develop in response to the requirements and feedback of users, who are provided with an initial, simplified working version of the model at the earliest feasible opportunity. -->

###	Assessment criteria
For each model attribute, we identified only assessment criteria that can be met with existing technical solutions and which therefore have the potential to be facilitated by software frameworks.

How **transparent** a CHEM is can be assessed against the criteria:

-	**T1**: All model code, non-confidential data and testing procedures and outcomes are available in open access repositories.

-	**T2**: It is easy to see who developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.

How **reusable** a CHEM is can be assessed against the criteria:

-	**R1**: Model code and data are implemented to facilitate both generalisability and transferability.

-	**R2**: Terms of use allow anyone to reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.

How **updatable** a CHEM is can be assessed against the criteria:

-	**U1**: Technical infrastructure is in place to support version control and collaboration with model users in the maintenance of a model.

-	**U2**: Each new release of a model is retested, with changes implemented to minimize disruptions for existing model users.

### 2023-08-15 Markdown/Manuscript/Child_RMDs/MS_Intro_Short.Rmd
working way through frameworks

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services [@wagstaff2012four].  In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and computer code (a computational model).  Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by applying the computational model. The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. Independent assessment of the validity of this assumption depends in part on the approach taken to computational model implementation and dissemination.

@@

Computational health economic models (CHEMs) can be implemented using specialized commercial software or authored as software development projects in an open-source programming language such as Python [@python2009] or R [@RCORE2022]. Advantages of commercial modelling tools include simplicity and ease of use, but a software development approach may facilitate computational models that are more transparent, reusable and adaptable [@incerti2019r]. The most appropriate approach for a specific CHEM project may depend on both the desired features of the model being developed and the resource and time constraints within which the project needs to be delivered (see Table \@ref(tab:proscons)). Although health economists report strong interest in making CHEMs more transparent and accessible [@Pouwels2022], open-source CHEMs remain rare [@Feenstra2022; @Emerson2019].

@@

<!-- We are developing a model to explore multiple economic questions relating to the mental health of young people aged 12 to 25.   -->
<!-- When making choices about how to implement this health economic model computationally, we wished to facilitate ethical development and use. -->
A high-level software framework using R has previously been developed with the primary aim of improving the transparency of CHEMs [@alarid2019need]. Novel software frameworks may be necessary to facilitate CHEM implementations that meet broader ethical criteria. The ethical responsibilities of computational modellers are often poorly understood and inadequately fulfilled [@10.3389/fpubh.2017.00068; @pliakos2021ethics; @thompson2022escape; @calder2018computational]. Guidance on developing, selecting and using software frameworks for ethical CHEM implementations is scarce.

>>>>

Health economics is a discipline concerned with problems that arise due to scarce resources, such as how to value health and healthcare, allocate healthcare budgets and configure health services [@wagstaff2012four].  In seeking to solve these problems, health economists typically use models which are simplified and selective representations of systems that are believed to influence human health. A health economic model should be capable of being described using words and figures (a conceptual model), equations (a mathematical model) and computer code (a computational model).  Health economic scientific manuscripts typically describe a model using its conceptual and mathematical representations but report results that have been generated by its computational representation (i.e., through execution of a computer program that applies the computational model to selected input data). The conceptual, mathematical and computational representations of a model are assumed to be isomorphic. Independent assessment of the validity of this assumption depends in part on how a computational model has been authored and shared.

@@

Computational health economic models (CHEMs) can be implemented using specialized commercial software or authored as software development projects in an open-source programming language such as Python [@python2009] or R [@RCORE2022]. Advantages of commercial modelling tools include simplicity and ease of use, but an open-source software development approach may facilitate computational models that are more transparent, reusable and updatable [@incerti2019r; @Pouwels2022]. The most appropriate implementation choice for a specific CHEM project may depend on both desired model features and the resource and time constraints within which the project needs to be delivered (see Table \@ref(tab:proscons)).

@@

<!-- We are developing a model to explore multiple economic questions relating to the mental health of young people aged 12 to 25.   -->
<!-- When making choices about how to implement this health economic model computationally, we wished to facilitate ethical development and use. -->
A high-level software framework for implementing open-source CHEMs in R has previously been developed with the primary aim of improving model transparency [@alarid2019need]. However, open-source CHEMs remain rare [@Feenstra2022; @Emerson2019], and novel software frameworks may be necessary help health economists fulfill a broader set of ethical modelling practice requirements. The ethical responsibilities of computational modellers are often poorly understood and inadequately fulfilled [@10.3389/fpubh.2017.00068; @pliakos2021ethics; @thompson2022escape; @calder2018computational]. Ideally, a novel software framework for CHEMs would also address some of the barriers preventing health economists adopting open-source approaches. The main barriers identified in a 2020 survey of health economists concern updating models, legal considerations, transferring model data, the platforms or languages in which models are developed, the level of public access required, lack of interest from decision-makers, model generalisability and confidentiality and security considerations [@Pouwels2022].

<!-- Guidance on developing, selecting and using software frameworks for ethical CHEM implementations is scarce. -->
